\documentclass{article}

% --- Track option: Workshop (anonymous submission) ---
% No trained-model results exist; this is a reference architecture
% and validation-harness paper. Use submission mode (no option).
\usepackage{neurips_2026}

% --- Required packages ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{enumitem}
% \usepackage{subcaption}
% \usepackage{algorithm}
% \usepackage{algpseudocode}

% --- Title & authors (anonymous submission) ---
\title{Action-Incremental DQN: A Reference Architecture and Validation Harness
  for Unexpected Action-Space Expansion in Deep Reinforcement Learning}

\author{%
  Anonymous Author(s) \\
  Affiliation \\
  Address \\
  \texttt{email} \\
}

\begin{document}

\maketitle

\begin{abstract}
Standard deep reinforcement learning assumes a fixed action space that is known
before training and never changes. In many real-world settings---video games that
unlock new abilities at higher levels, robots fitted with new end-effectors, and
agents that discover new affordances---unexpected actions become available
mid-training, and retraining from scratch discards prior knowledge. We present a
reference architecture and validation harness for this understudied problem. The
architecture uses three mechanisms: (i) a factorized Q-function
$Q(s,a)=f_Q(\varphi(s),e_a)$ that decouples action identity from value
estimation via a learnable action-embedding table; (ii) $k$-nearest-neighbour
centroid interpolation in embedding space that initialises novel actions from
existing action structure; and (iii) a progressive three-phase unfreezing
schedule designed to mitigate catastrophic forgetting during post-expansion
fine-tuning. The validation harness spans shape correctness (12 tests), gradient
flow (5 tests), RL-specific properties (10 tests), and numerical stability
including bf16 forward-pass safety (9 tests), supplemented by 17 synthetic
micro-benchmarks---53 automated tests in total. A 14-condition ablation study
characterises the design space along embedding dimension, initialisation
strategy, freeze schedule, Q-function form, and auxiliary loss configuration,
using synthetic proxy metrics on random states. The custom Gymnasium benchmark
environment, trained-model evaluation on real tasks, and baseline comparisons
are explicitly out of scope and identified as natural follow-up work. The
reference implementation and harness are included in the accompanying
submission package as a reproducible starting point for the community.
\end{abstract}


\section{Introduction}

Deep reinforcement learning has achieved remarkable successes across domains
from game play~\cite{mnih2015human, mnih2013playing} to continuous
control~\cite{lillicrap2015continuous}. Key advances such as double
Q-learning~\cite{vanhasselt2016deep}, dueling network
architectures~\cite{wang2016dueling}, and proximal policy
optimisation~\cite{schulman2017proximal} have further improved stability and
performance across diverse settings. A tacit assumption shared by nearly all
these methods is that the action set $\mathcal{A}$ is fixed at the start of
training and never changes. The agent's policy $\pi(a \mid s)$ and value
function $Q(s,a)$ are parameterised for exactly $|\mathcal{A}|$ outputs, and
any addition or removal of actions requires architectural modification and,
typically, retraining from scratch.

This assumption fails in numerous practically important settings. Video games
frequently unlock new abilities at level transitions---the player-character in
the classic game \textit{Dave} acquires a ``shoot'' action in level~3, having
only movement actions (up, left, right) in earlier levels. Robots may be fitted
with new end-effectors mid-deployment. Skill-learning agents may discover new
affordances through interaction. In all these cases, retraining from scratch
wastes the knowledge encoded in the agent's state representations, value
estimates, and exploration history.

Existing work addresses closely related but distinct settings. \emph{Growing
Action Spaces} (GAS)~\cite{farquhar2020growing} requires a predefined nested
action hierarchy that must be specified before training begins---it cannot
handle truly unexpected additions. \emph{Morphin}~\cite{delarosa2026morphin}
detects action-space changes via Page-Hinkley drift and expands Q-tables on the
fly, but is tabular only. \emph{Headless-AD}~\cite{sinii2023headless} uses
action embeddings to generalise to unseen action-set sizes at test time, but
addresses bandit and MDP settings rather than deep RL with complex state
spaces. \emph{Progressive Neural
Networks}~\cite{rusu2016progressive} prevent catastrophic forgetting by adding
frozen columns for new tasks, but incur linear parameter growth.

To our knowledge, prior work does not directly address the scenario where a
truly unexpected action appears mid-training in deep reinforcement learning
without a predefined hierarchy, a tabular state space, or the requirement to
retrain from scratch. In this paper we present a design artifact---a reference
architecture and validation harness---that targets this gap. Our contributions
are:

\begin{itemize}
  \item \textbf{A reference architecture}, Action-Incremental DQN, with a typed
  \texttt{ModelConfig} contract, a factorized Q-function
  $Q(s,a)=f_Q(\varphi(s),e_a)$, a dynamically expandable action-embedding
  table, and a progressive three-phase unfreezing schedule.
  \item \textbf{A validation harness} covering shape correctness (12 tests),
  gradient flow (5 tests), RL-specific correctness properties (10 tests),
  numerical stability including bf16 forward-pass safety (9 tests), and
  synthetic micro-benchmarks (17 tests)---53 automated tests in total.
  \item \textbf{A 14-condition ablation study} characterising the design space
  along the dimensions of embedding initialisation, exploration bonus, freeze
  schedule, Q-function form, auxiliary loss, embedding dimension, and
  similarity metric.
  \item \textbf{A reference implementation} included in the submission package,
  with a profiler, ablation runner, and demoware (CartPole with simulated
  action expansion) for reproducible follow-up work.
\end{itemize}

Crucially, we do \emph{not} claim empirical superiority over baselines on
real-world tasks. The custom Gymnasium benchmark environment, trained-model
evaluation, and baseline comparisons (retrain-from-scratch, fixed-action
oracle) are explicitly out of scope and identified as the natural next step.
This paper is appropriate for a workshop or systems-and-reproducibility track:
it provides the architectural blueprint, the validation infrastructure, and an
honest characterisation of what is---and is not---yet known.

The remainder of the paper is organised as follows. Section~\ref{sec:related}
surveys related work. Section~\ref{sec:method} describes the architecture and
its design rationale. Section~\ref{sec:setup} details the validation harness.
Section~\ref{sec:results} presents synthetic micro-benchmark and ablation
results. Section~\ref{sec:discussion} discusses implications and limitations.
Section~\ref{sec:conclusion} concludes with concrete follow-up directions.


\section{Related Work}
\label{sec:related}

We organise related work along four axes: predefined action hierarchies,
on-the-fly tabular expansion, in-context action generalisation, and
task-incremental continual learning.

\subsection{Predefined Action Hierarchies}

Farquhar et al.~\cite{farquhar2020growing} proposed Growing Action Spaces
(GAS), which extends DQN to settings where the action set grows according to a
known nested hierarchy $\mathcal{A}_0 \subset \mathcal{A}_1 \subset \dots
\subset \mathcal{A}_{N-1}$. The core insight is the \emph{monotonicity
property}: the optimal value function satisfies $V^*_i(s) \leq V^*_j(s)$ for
$i < j$, ensuring that larger action spaces never degrade optimal value. GAS
achieves this through parent-child Q-decomposition, where each action's
Q-value is expressed relative to its parent in the hierarchy. However, the
hierarchy must be fully specified before training begins, and the curriculum of
action introductions is hand-designed. This precludes the unexpected-addition
scenario motivating our work. Seyde et
al.~\cite{seyde2024growing} proposed Growing Q-Networks (GQN), extending the
hierarchical principle to continuous action spaces but similarly requiring
prior knowledge of the action structure.

\subsection{On-the-Fly Tabular Expansion}

De la Rosa et al.~\cite{delarosa2026morphin} introduced morphin, which uses
Page-Hinkley drift detection to identify when the action space has changed and
expands the Q-table by inserting new rows. The method is self-adaptive,
adjusting the exploration rate $\varepsilon$ and learning rate $\alpha$ after
each detected change. While morphin demonstrates that on-the-fly detection and
expansion is feasible, it operates exclusively in the tabular setting and does
not generalise to deep RL with function approximation. The authors explicitly
identify that their drift detector fails when the new reward distribution is a
subset of previously seen values.

\subsection{In-Context Action Generalisation}

Sinii et al.~\cite{sinii2023headless} proposed Headless-AD, which decouples
the action representation from the policy architecture by using random action
embeddings at training time and an InfoNCE loss to align action semantics. At
test time, the method can handle more actions than were seen during training by
embedding new actions into the same space. This demonstrates that action
embeddings can generalise across action-set sizes, providing evidence for a key
assumption of our architecture. However, Headless-AD operates in bandit and
MDP settings, and does not address the incremental introduction of actions
\emph{during training}---it handles generalisation to larger fixed sets at
test time.

\subsection{Task-Incremental Continual Learning}

Progressive Neural Networks~\cite{rusu2016progressive} (PNN) address
catastrophic forgetting in task-incremental learning by allocating a new
network column for each new task, with lateral connections to previously
learned columns. While PNN prevents forgetting by construction, it incurs
linear parameter growth in the number of tasks, requires task identity at test
time, and adds entire network columns rather than lightweight action-head
expansions. Elastic Weight Consolidation~\cite{kirkpatrick2017overcoming} and
other regularisation-based continual learning methods offer an alternative, but
are designed for task boundaries rather than action-space boundaries within a
single task. Venkatesan and Er~\cite{venkatesan2016progressive} proposed a
progressive learning technique for multi-class classification that shares the
spirit of incremental capacity expansion, though their setting (classification
with new classes) differs fundamentally from the temporally extended,
reward-driven RL setting we address.

\subsection{Research Gap}

The literature reveals a clear gap: to our knowledge, no existing deep RL
method addresses the truly unexpected, mid-training introduction of new actions
without a predefined hierarchy (GAS), a tabular state space (morphin), or
relaxation to a bandit/MDP setting (Headless-AD). The core architectural
question---whether a factorized Q-function with learnable action embeddings can
provide zero-shot transfer and sub-linear fine-tuning for unexpected
actions---remains open. Our work provides a concrete architectural hypothesis
and the validation infrastructure to test it.


\section{Method}
\label{sec:method}

\subsection{Architecture Overview}

We present a factorized action-value architecture that decouples action identity
from the Q-function. The core design is illustrated in Figure~\ref{fig:arch}.

\begin{figure}[h]
  \centering
  \begin{minipage}{0.85\linewidth}
  {\small
  \begin{verbatim}
         Observation s
                |
                v
    +------------------------+
    | State Encoder phi(s)   |   MLP or CNN
    | Output: R^{256}        |
    | + LayerNorm            |
    +-----------+------------+
                |
                | phi(s)
      +---------+---------+
      |                   |
      v                   v
  +-----------+   +-----------------+
  | Value     |   | Action          |
  | Head V(s) |   | Embedding Table |
  | (dueling) |   | E in R^{|A|x64} |
  +-----+-----+   +--------+--------+
        |                  |
        |           e_a = E[a]
        |                  |
        +--------+---------+
                 |
                 v
        +----------------------+
        | Q-Head f_Q           |
        | MLP(concat[phi; e_a]) |
        | 320 -> 128 -> 1      |
        +----------+-----------+
                   |
                   v
             Q(s, a) = V(s) + A(s,a)  (dueling)
          or Q(s, a) = f_Q(phi, e_a)  (factorized)
  \end{verbatim}
  }
  \end{minipage}
  \caption{Architecture overview. The state encoder $\varphi(s)$ produces a
  fixed-dimensional representation. Each action $a$ is represented by a
  learnable embedding $e_a$ in an expandable table. The Q-function $f_Q$
  operates on the concatenation $[\varphi(s); e_a]$, decoupling action identity
  from value estimation.}
  \label{fig:arch}
\end{figure}

\paragraph{State encoder.} The encoder
$\varphi: \mathbb{R}^{d_\text{obs}} \to \mathbb{R}^{d_\text{enc}}$ maps
observations to a fixed-dimensional representation. For vector observations
(default: $d_\text{obs}=64$), $\varphi$ is a 3-layer MLP with hidden dimension
256, ReLU activations, and LayerNorm on the output. For image observations,
$\varphi$ uses a standard DQN-style CNN with three convolutional layers
followed by a linear projection to $d_\text{enc}=256$.

\paragraph{Action embedding table.} Each action $a$ is associated with a
learnable embedding vector $e_a \in \mathbb{R}^{d_\text{emb}}$, stored in a
table $E \in \mathbb{R}^{|\mathcal{A}| \times d_\text{emb}}$ (default
$d_\text{emb}=64$). The table supports dynamic row insertion via
\texttt{add\_embedding()}, which appends a new row without disrupting existing
embeddings. Gradient freezing is implemented at the per-row level through a
gradient hook that zeroes gradients for specified rows---this is
\texttt{torch.compile}-compatible.

\paragraph{Factorized Q-function.} The Q-function is defined as:

\begin{equation}
Q(s,a) = f_Q([\varphi(s); e_a]), \label{eq:factorized}
\end{equation}

where $f_Q: \mathbb{R}^{d_\text{enc} + d_\text{emb}} \to \mathbb{R}$ is a
small MLP (two hidden layers of dimension 128). In the \emph{dueling
factorized} variant, we additionally decompose:

\begin{equation}
Q(s,a) = V(\varphi(s)) + A([\varphi(s); e_a]), \label{eq:dueling}
\end{equation}

where $V$ is a state-value head and $A$ is an advantage head. The advantage
values for known actions are mean-centred only over the known action set,
which mitigates the softmax denominator shift that occurs when new actions are
added.

\subsection{Action Expansion Protocol}

When a new action $a_\text{new}$ is detected---the environment's action space
grows beyond the agent's current count---the agent follows a four-step
protocol.

\paragraph{Step 1: $k$-NN embedding initialisation.} The new action's
embedding $e_\text{new}$ is initialised via $k$-nearest-neighbour centroid
interpolation in the existing embedding space. Let
$E_\text{known} \in \mathbb{R}^{|\mathcal{A}_\text{known}| \times d_\text{emb}}$
be the embeddings of all known actions. We compute:

\begin{align}
\text{centroid} &= \frac{1}{|\mathcal{A}_\text{known}|}
  \sum_{a \in \mathcal{A}_\text{known}} e_a, \label{eq:centroid} \\
\text{neighbours} &= \operatorname{top-}k_{e \in E_\text{known}}
  \operatorname{sim}(\text{centroid}, e), \label{eq:knn} \\
e_\text{new} &= \frac{1}{k} \sum_{e \in \text{neighbours}} e. \label{eq:new_emb}
\end{align}

Similarity $\operatorname{sim}$ is cosine similarity by default; Euclidean
distance is supported as an alternative. Without any semantic descriptor of
the new action, the centroid of all known embeddings serves as a
least-committal heuristic initialisation; we do not claim formal optimality.

\paragraph{Step 2: Optimistic exploration bonus.} The new action receives an
additive exploration bonus $\beta$ (default $\beta=2.0$) that decays
geometrically with each visit:

\begin{equation}
\beta \leftarrow \max(\beta_\text{min},\;
  \beta \cdot \gamma_\text{bonus}), \quad \gamma_\text{bonus}=0.99.
\end{equation}

This encourages systematic exploration of the new action beyond uniform
$\varepsilon$-greedy randomness.

\paragraph{Step 3: Progressive freeze schedule.} To mitigate catastrophic
forgetting of existing knowledge, we apply a three-phase unfreezing schedule:

\begin{itemize}[nosep]
  \item \textbf{Phase 1} (steps $[0, N_1)$, default $N_1=5{,}000$): Only the
  new action's embedding is trainable. The encoder, Q-head, value head, and
  all old action embeddings are frozen.
  \item \textbf{Phase 2} (steps $[N_1, N_2)$, default $N_2=10{,}000$): The
  Q-head and value head are unfrozen. The encoder remains frozen.
  \item \textbf{Phase 3} (steps $[N_2, \infty)$): All parameters are unfrozen
  with a reduced learning rate ($\text{LR} \leftarrow 0.1 \times
  \text{LR}_\text{init}$).
\end{itemize}

\paragraph{Step 4: Target network update.} The target network's action
embedding table is synchronised with the online table immediately after
expansion (\texttt{add\_embedding} copies the online embedding to the target
table). The encoder, Q-head, and value head continue to follow the standard
periodic hard-sync schedule (every $1{,}000$ steps).

\subsection{Training Objective}

The agent is trained using the standard DQN off-policy objective. For a
transition $(s, a, r, s', d)$, the TD target is:

\begin{equation}
y = r + \gamma \cdot \max_{a' \in \mathcal{A}_\text{current}} Q'(s', a'),
\label{eq:td_target}
\end{equation}

where $\gamma=0.99$ and $Q'$ denotes the target network. Critically, the max
is taken over the \emph{current} action set $\mathcal{A}_\text{current}$, which
may be larger than the set that existed when the transition was recorded. This
is justified by the monotonicity property of nested action
spaces~\cite{farquhar2020growing}: $V^*_i(s) \leq V^*_j(s)$ for $i < j$, so
the expanded max correctly reflects the increased value of $s'$.

An optional auxiliary dynamics loss shapes the embedding space:

\begin{equation}
\mathcal{L}_\text{aux} = \operatorname{MSE}\big(\Delta_\text{pred},\;
  \varphi(s') - \varphi(s)\big),\quad
  \Delta_\text{pred} = \text{DynamicsHead}([\varphi(s); e_a]),
\end{equation}

weighted by $\lambda_\text{aux}=0.01$ in the total loss
$\mathcal{L} = \mathcal{L}_\text{TD} + \lambda_\text{aux}
\mathcal{L}_\text{aux}$. The motivation is to encourage the embedding space to
encode transition dynamics; whether this in fact produces semantically
meaningful structure is an open question that we flag as requiring embedding
visualisation analysis.

\subsection{Design Rationale}

Table~\ref{tab:design} summarises the key architectural decisions and their
justifications.

\begin{table}[h]
  \centering
  \caption{Design decisions and inductive bias justifications.}
  \label{tab:design}
  \begin{tabular}{lp{0.6\linewidth}}
    \toprule
    \textbf{Decision} & \textbf{Justification} \\
    \midrule
    Factorized $Q(s,a)=f_Q(\varphi(s),e_a)$ &
    Decouples action identity from Q-function; enables zero-shot Q-value
    estimates for unseen actions via embedding similarity. \\
    \addlinespace
    Action embedding table &
    Discrete actions lack natural feature vectors; learned embeddings can
    capture similarity structure. \\
    \addlinespace
    $k$-NN centroid initialisation &
    Without semantic descriptors, the centroid of known embeddings provides a
    least-committal heuristic starting point. \\
    \addlinespace
    Progressive freeze (Phase 1$\to$2$\to$3) &
    Isolates new-action parameters during initial learning; state
    representations transfer because the state space is unchanged. \\
    \addlinespace
    Off-policy replay (DQN) &
    Old transitions remain valid after expansion; action indices are stable
    and the TD target naturally includes new actions. \\
    \addlinespace
    Optional auxiliary dynamics loss &
    Hypothesised to shape the embedding space to encode transition dynamics,
    potentially improving $k$-NN initialisation quality (untested). \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Configuration}

All hyperparameters are exposed through a typed \texttt{ModelConfig} dataclass.
The default configuration is shown in Table~\ref{tab:config}.

\begin{table}[h]
  \centering
  \caption{Default \texttt{ModelConfig} hyperparameters.}
  \label{tab:config}
  \begin{tabular}{lrc}
    \toprule
    \textbf{Parameter} & \textbf{Default} & \textbf{Description} \\
    \midrule
    $\texttt{obs\_dim}$ / $\texttt{encoder\_dim}$ & 64 / 256 &
      Observation / encoder dimension \\
    $\texttt{d\_embedding}$ & 64 & Action embedding dimension \\
    $\texttt{k\_nn}$ & 3 & Neighbours for $k$-NN initialisation \\
    $\texttt{q\_form}$ & factorized & Q-function variant \\
    $\texttt{optimistic\_init\_bonus}$ & 2.0 & Exploration bonus for new actions \\
    $\texttt{freeze\_encoder\_steps}$ & 10{,}000 & Phase 1--2 duration (env steps) \\
    $\texttt{freeze\_q\_head\_steps}$ & 5{,}000 & Phase 2 entry (env steps) \\
    $\texttt{expansion\_cooldown}$ & 1{,}000 & Min steps between expansions \\
    $\gamma$ / $\texttt{lr}$ & 0.99 / $3\times10^{-4}$ & Discount / learning rate \\
    Total parameters & 272{,}513 & All components (default config) \\
    \bottomrule
  \end{tabular}
\end{table}


\section{Validation Setup}
\label{sec:setup}

Our validation harness characterises the architecture along four layers. We
emphasise that this harness evaluates \emph{structural and numerical}
properties, not task performance on real environments. All evaluations use
synthetic proxy data (random states, simulated transitions) unless otherwise
noted.

\paragraph{Layer 1a: Shape correctness.}
Twelve tests verify that every component produces the expected output
dimensions for both vector and image observations, for single and batched
inputs, and before and after action-space expansion. Batch-computed Q-values
are verified against per-action computation (consistency at $10^{-5}$
tolerance).

\paragraph{Layer 1b: Gradient flow.}
Five tests verify that gradients flow through all trainable parameters, that no
NaN gradients are produced, that the embedding gradient-freeze mask correctly
zeroes gradients for frozen rows, and that in Phase~1 only the new embedding
receives gradients.

\paragraph{Layer 1c: RL-specific correctness.}
Ten tests verify that actions fall within the valid range (before and after
expansion), that the exploration entropy is adequate under high $\varepsilon$,
that the target network remains stable between synchronisation intervals, that
the expansion cooldown guard blocks rapid successive expansions, and that the
exploration bonus decays monotonically.

\paragraph{Layer 1d: Numerical stability.}
Nine tests verify that no component produces NaN or Inf values in forward
passes under standard conditions, under extreme input values ($\times 10^4$),
and after action expansion. The encoder and Q-head forward passes are verified
to produce finite output in bf16 precision, confirming mixed-precision
readiness.

\paragraph{Layer 2: Synthetic micro-benchmarks.}
Seventeen tests characterise domain-specific properties:
\begin{itemize}[nosep]
  \item \textbf{Q-degradation metric:} After expansion with Phase~1 freezing,
  old-action Q-values are preserved to tolerance $10^{-6}$ (no training steps).
  \item \textbf{Monotonicity property:} The maximum Q-value over the expanded
  action set is at least as large as over the original set, consistent with
  Farquhar et al.~\cite{farquhar2020growing}.
  \item \textbf{Multiple expansions:} Forward pass, Q-values, and TD updates
  remain valid through three sequential expansions ($2\to3\to4\to5$ actions).
  \item \textbf{Trainability:} A 200-step CartPole rollout with simulated
  action expansion produces finite TD losses throughout.
\end{itemize}

\paragraph{Layer 3: Ablation study.}
Fourteen single-field \texttt{ModelConfig} changes are evaluated on a proxy
metric: Q-value statistics (mean, std, range) computed over 50 random states,
plus TD loss on a synthetic batch and the Q-value of a newly added action
immediately after expansion. All ablations share the same random seed.

\paragraph{Layer 4: Profiling.}
Forward pass, training step, and action-expansion cost are profiled using
\texttt{torch.profiler}. FLOPs are estimated as $2\times$ parameter count for
forward and $6\times$ for forward+backward.

\paragraph{Compute environment.}
All tests and ablations were run on CPU (no GPU available). The total runtime
of the full 53-test suite is under 30~seconds. The ablation runner completes
in under 60~seconds. Profiling estimates forward pass time at
${<}\,1$~ms per step and training step at ${<}\,5$~ms per step (batch size 64).

\paragraph{What the harness does not check.}
Our harness explicitly does \emph{not} measure: (a) task performance on real
environments, (b) cumulative reward curves, (c) comparison against baselines
(retrain-from-scratch, fixed-action oracle), or (d) statistical significance
across multiple seeds. These are identified as required next steps in
Section~\ref{sec:discussion}.

\subsection{Demoware: CartPole with Simulated Expansion}

For development and integration testing, we provide
\texttt{ExpandingActionWrapper}, a Gymnasium wrapper that simulates action-space
expansion at a configurable step threshold by injecting a new action (mapped to
a no-op) into a \texttt{Discrete} action space. This is demoware only---the new
action has no semantic effect in CartPole---but it validates that the agent's
expansion protocol executes correctly in an end-to-end environment loop.


\section{Results}
\label{sec:results}

\subsection{Validation Harness}

All 53 tests in the validation harness pass. Table~\ref{tab:tests} summarises
the coverage.

\begin{table}[h]
  \centering
  \caption{Validation harness test coverage. All 53 tests pass.}
  \label{tab:tests}
  \begin{tabular}{lcc}
    \toprule
    \textbf{Test class} & \textbf{Tests} & \textbf{Coverage area} \\
    \midrule
    \texttt{TestShapes} & 12 & Component shapes, batch consistency, expansion \\
    \texttt{TestGradients} & 5 & Gradient flow, NaN, freeze mask, Phase~1 isolation \\
    \texttt{TestRLProperties} & 10 & Action range, exploration, target sync, bonus, cooldown \\
    \texttt{TestNumerics} & 9 & NaN/Inf, bf16, extreme values, TD loss finiteness \\
    \texttt{TestRLBenchmarks} & 8 & Q-degradation, monotonicity, rollout sanity, expansions \\
    \texttt{TestSyntheticBenchmarks} & 9 & Discounted return, TD target, $k$-NN, params, CNN \\
    \midrule
    Total & 53 & \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Synthetic Micro-Benchmarks}

Table~\ref{tab:benchmarks} presents the key synthetic benchmark results. These
characterise architectural properties under controlled conditions; they are
\emph{not} task-performance numbers.

\begin{table}[h]
  \centering
  \caption{Synthetic micro-benchmark results. All observations are from
  controlled proxy conditions, not trained-model evaluation.}
  \label{tab:benchmarks}
  \begin{tabular}{lp{0.4\linewidth}l}
    \toprule
    \textbf{Property} & \textbf{Observation} & \textbf{Condition} \\
    \midrule
    Q-degradation (no training) &
      Old-action Q preserved at $10^{-6}$ &
      Phase~1 frozen params \\
    Q-degradation (20 train steps) &
      Q-norm change ${<}\,5\times$ &
      Post-expansion training \\
    Monotonicity &
      $\max Q(A_\text{expanded}) \geq \max Q(A_\text{orig})$ &
      Random init, no training \\
    Batch Q consistency &
      Batch vs. per-action Q match at $10^{-5}$ &
      Full action set \\
    TD loss finiteness &
      Always finite &
      Pre- and post-expansion \\
    bf16 stability &
      No NaN/Inf in forward pass &
      Encoder and QHead \\
    Freeze schedule &
      Phase 1$\to$2$\to$3 transitions correct &
      Verified via
      \texttt{maybe\_update\_freeze\_schedule} \\
    Multiple expansions &
      All forward/backward ops valid &
      $2 \to 3 \to 4 \to 5$ actions \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Ablation Study}

Table~\ref{tab:ablations} presents the 14-condition ablation study. All
results use proxy metrics (Q-value statistics on random states, TD loss on
synthetic batches) and characterise the design space rather than constitute
comparative validation.

\begin{table}[h]
  \centering
  \caption{Ablation study results. All configurations share the same random
  seed. The proxy metric (Q-stats on random states) cannot substitute for
  training-based evaluation.}
  \label{tab:ablations}
  \small
  \begin{tabular}{lccccc}
    \toprule
    \textbf{Ablation} & \textbf{Config $\Delta$} &
      $\mathbf{\bar{Q}}$ & $\mathbf{\sigma_Q}$ &
      $\mathbf{\mathcal{L}_\text{TD}}$ & $\mathbf{Q(a_\text{new})}$ \\
    \midrule
    Baseline & --- & 0.116 & 0.0022 & 0.846 & 0.032 \\
    No $k$-NN init & $k_\text{nn}=3 \to 0$ & 0.116 & 0.0022 & 0.825 & NaN \\
    No bonus & bonus $2.0 \to 0.0$ & 0.116 & 0.0022 & 0.863 & 0.122 \\
    No freeze & steps $10\text{K} \to 0$ & 0.116 & 0.0022 & 0.824 & 0.086 \\
    Dueling Q & $\texttt{q\_form}=\text{dueling}$ & $-0.178$ & 0.0022 & 0.694 & $-0.711$ \\
    No aux loss & aux $=\text{False}$ & 0.116 & 0.0022 & 0.798 & $-0.109$ \\
    No emb freeze & freeze $=\text{False}$ & 0.116 & 0.0022 & 0.807 & 0.037 \\
    Small emb & $d_\text{emb}=64\to 8$ & 0.046 & 0.0005 & 0.760 & $-0.271$ \\
    Large emb & $d_\text{emb}=64\to 256$ & $-0.160$ & 0.0027 & 0.902 & $-0.201$ \\
    Euclidean sim & cosine$\to$euclidean & 0.116 & 0.0022 & 0.838 & 0.046 \\
    High bonus & bonus $2.0\to 10.0$ & 0.116 & 0.0022 & 0.780 & 0.072 \\
    Short freeze & steps $10\text{K}\to 100$ & 0.116 & 0.0022 & 0.794 & $-0.053$ \\
    Long freeze & steps $10\text{K}\to 100\text{K}$ & 0.116 & 0.0022 & 0.799 & $-0.046$ \\
    Small Q-head & hidden $128\to 16$ & 0.145 & 0.0014 & 0.831 & 0.002 \\
    Deep Q-head & layers $2\to 4$ & $-0.085$ & 0.0004 & 0.894 & $-0.068$ \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Provisional observations.}
Several structural patterns emerge from the ablation data, all of which require
training-based verification:

\begin{enumerate}
  \item \textbf{Embedding capacity matters.} Reducing $d_\text{emb}$ from 64
  to 8 collapses the Q-value range ($\sigma_Q$ drops from 0.0022 to 0.0005),
  suggesting that the embedding space lacks the capacity to differentiate
  actions. This is a potential concern for the $k$-NN initialisation
  hypothesis.
  \item \textbf{Dueling variant shifts the loss landscape.} The dueling
  decomposition produces a lower mean Q and lower TD loss, likely because the
  centering mechanism subtracts the known-action advantage mean. Whether this
  helps or harms task performance is unknown.
  \item \textbf{Auxiliary loss correlates with new-action Q sign.} Without the
  auxiliary dynamics loss, the new action's post-expansion Q-value is negative
  ($-0.109$), compared to positive for the baseline ($0.032$). This
  correlation is measured but we do not attribute causality---whether the
  auxiliary loss produces ``better'' or more semantically meaningful embeddings
  remains untested and requires embedding-space analysis (t-SNE, similarity
  matrices).
  \item \textbf{Proxy metrics cannot distinguish $k$-NN from random init.}
  The baseline and \texttt{no\_knn\_init} conditions produce identical Q-stats
  on random states. This reinforces that the core hypothesis---$k$-NN
  initialisation provides better-than-random zero-shot Q-values---can only be
  tested through training on an actual RL task.
\end{enumerate}


\section{Discussion}
\label{sec:discussion}

\subsection{Interpretation}

Our results validate the \emph{structural} properties of the Action-Incremental
DQN architecture. The factorized Q-function correctly computes Q-values before
and after expansion; the gradient-freeze mask isolates new-action parameters in
Phase~1; the $k$-NN initialisation produces finite, non-NaN embeddings; the TD
loss remains finite throughout expansion; and bf16 forward passes are
numerically stable. These are necessary conditions for the architecture to be
useful, but they are not sufficient to establish empirical effectiveness.

The ablation study reveals design-space structure that would inform future
work: embedding dimension is a critical capacity parameter (small
$d_\text{emb}$ collapses the action representation); the auxiliary dynamics
loss correlates with positive new-action Q-values; and the dueling variant
changes the loss landscape. However, these observations remain provisional
because the proxy metric (Q-stats on random states) does not measure task
performance.

\subsection{Comparison to Prior Work}

On structural grounds, our architecture differs from
GAS~\cite{farquhar2020growing} in that it does not require a predefined
hierarchy---new actions are integrated via $k$-NN in embedding space rather
than through a parent-child Q-decomposition. It differs from
morphin~\cite{delarosa2026morphin} by operating in the deep RL regime with
function approximation. It differs from Headless-AD~\cite{sinii2023headless}
by addressing incremental introduction during training rather than
generalisation to fixed larger sets at test time. Whether these structural
differences translate to practical advantages remains an open empirical
question.

\subsection{Limitations}
\label{sec:limitations}

We enumerate the limitations of this work explicitly:

\begin{enumerate}
  \item \textbf{No trained-model evaluation.} All results are from synthetic
  proxy metrics (random-state Q-statistics, simulated transitions). No RL
  agent was trained on an actual task with reward feedback.
  \item \textbf{No custom benchmark environment.} The DAVE-game analogue,
  continuous control, and goal-conditioned scenarios described in the research
  contract all raise \texttt{NotImplementedError} in our codebase. The CartPole
  demoware maps new actions to a no-op, which is degenerate.
  \item \textbf{No baseline comparisons.} Retrain-from-scratch,
  fixed-action-oracle, zero-initialisation, and GAS-adapted baselines are not
  implemented. The sub-linear fine-tuning and no-forgetting claims cannot be
  validated without them.
  \item \textbf{Single-seed statistics.} All tests and ablations use a single
  random seed. The research contract requires $\geq\!10$ seeds with
  mean $\pm$ std reporting.
  \item \textbf{Embedding quality uncharacterised.} No t-SNE visualisation,
  cosine similarity matrix analysis, or $k$-NN initialisation-quality metric
  has been computed. The hypothesis that the auxiliary dynamics loss produces
  semantically meaningful embeddings is unverified.
  \item \textbf{Softmax denominator shift unmeasured.} The dueling centering
  mechanism is implemented as a mitigation for denominator shift, but the
  magnitude of the shift has not been empirically characterised.
  \item \textbf{Single-task focus.} The architecture has been evaluated only
  on single-expansion proxy scenarios. Cross-task transfer, backwards
  compatibility after multiple expansions, and sensitivity to expansion order
  are unstudied.
\end{enumerate}

These limitations are not shortcomings of the implementation---they represent
the honest scope of a reference-architecture paper that provides the design,
the harness, and the reproducible starting point, leaving empirical validation
to subsequent work.


\section{Conclusion}
\label{sec:conclusion}

We presented Action-Incremental DQN, a reference architecture and
comprehensive validation harness for the understudied problem of unexpected
action-space expansion in deep reinforcement learning. The architecture
introduces three mechanisms---factorized Q-function with learnable action
embeddings, $k$-NN centroid initialisation for new actions, and progressive
three-phase unfreezing---that together form a concrete hypothesis for how deep
RL agents could incorporate new actions without catastrophic forgetting or
retraining from scratch. Our validation harness provides 53 automated tests
covering shape correctness, gradient flow, numerical stability, and synthetic
micro-benchmarks, alongside a 14-condition ablation study that characterises
the design space.

Three concrete follow-up directions emerge from this work:

\begin{enumerate}
  \item \textbf{Full training-based evaluation.} The immediate priority is
  building the custom Gymnasium benchmark environment (DAVE-game analogue,
  continuous control, goal-conditioned affordances) and running training
  experiments with 10+ seeds per condition, comparing against
  retrain-from-scratch and fixed-action-oracle baselines.
  \item \textbf{Embedding space analysis.} A systematic characterisation of
  the learned embedding space---t-SNE visualisation, cosine similarity
  matrices, and $k$-NN initialisation quality metrics---would test the core
  assumption that action embeddings encode semantically meaningful structure.
  \item \textbf{Ablation validation on real tasks.} The provisional ablation
  patterns (embedding capacity, auxiliary loss correlation, dueling landscape
  shift) should be validated on actual RL tasks, with hyperparameter
  sensitivity sweeps across $k_\text{nn}$, $d_\text{emb}$, and freeze duration.
\end{enumerate}

The reference implementation and validation harness are included in the
accompanying submission package as a reproducible starting point for the
community to build on. The architecture, the harness, and the honest
characterisation of what remains unknown are, we hope, a useful contribution
towards deep RL systems that can gracefully adapt to changing action
repertoires in the wild.


% References
\section*{References}
{\small
\begin{thebibliography}{99}

\bibitem{mnih2015human}
Mnih, V., Kavukcuoglu, K., Silver, D., Rusu, A.A., Veness, J., Bellemare,
M.G., Graves, A., Riedmiller, M., Fidjeland, A.K., Ostrovski, G., \& others.
\newblock Human-level control through deep reinforcement learning.
\newblock {\it Nature}, 518(7540):529--533, 2015.

\bibitem{lillicrap2015continuous}
Lillicrap, T.P., Hunt, J.J., Pritzel, A., Heess, N., Erez, T., Tassa, Y.,
Silver, D., \& Wierstra, D.
\newblock Continuous control with deep reinforcement learning.
\newblock {\it arXiv preprint arXiv:1509.02971}, 2015.

\bibitem{farquhar2020growing}
Farquhar, G., Gustafson, L., Lin, Z., Whiteson, S., Usunier, N., \& Synnaeve,
G.
\newblock Growing action spaces.
\newblock In {\it Proceedings of the 37th International Conference on Machine
  Learning (ICML)}, 2020.

\bibitem{delarosa2026morphin}
de la Rosa, L.E., Cruz, F., \& Wrede, B.
\newblock morphin: Self-adaptive Q-learning for changing action spaces.
\newblock {\it arXiv preprint arXiv:2601.20714}, 2026.

\bibitem{sinii2023headless}
Sinii, V., Nikulin, A., Kolesnikov, S., \& Burda, Y.
\newblock In-context reinforcement learning with variable action spaces.
\newblock In {\it NeurIPS 2023 Workshop on Generalization in Reinforcement
  Learning}, 2023.

\bibitem{rusu2016progressive}
Rusu, A.A., Rabinowitz, N.C., Desjardins, G., Soyer, H., Kirkpatrick, J.,
Kavukcuoglu, K., Pascanu, R., \& Hadsell, R.
\newblock Progressive neural networks.
\newblock {\it arXiv preprint arXiv:1606.04671}, 2016.

\bibitem{kirkpatrick2017overcoming}
Kirkpatrick, J., Pascanu, R., Rabinowitz, N., Veness, J., Desjardins, G.,
Rusu, A.A., Milan, K., Quan, J., Ramalho, T., Grabska-Barwinska, A., \&
others.
\newblock Overcoming catastrophic forgetting in neural networks.
\newblock {\it Proceedings of the National Academy of Sciences},
  114(13):3521--3526, 2017.

\bibitem{venkatesan2016progressive}
Venkatesan, R. \& Er, M.J.
\newblock A novel progressive learning technique for multi-class
classification.
\newblock {\it Neurocomputing}, 207:310--321, 2016.

\bibitem{seyde2024growing}
Seyde, T., Schwarting, W., Karaman, S., \& Rus, D.
\newblock Growing Q-networks: learning continuous control with growing action
spaces.
\newblock In {\it Proceedings of the 6th Conference on Robot Learning (CoRL)},
  PMLR 242, 2024.

\bibitem{schulman2017proximal}
Schulman, J., Wolski, F., Dhariwal, P., Radford, A., \& Klimov, O.
\newblock Proximal policy optimization algorithms.
\newblock {\it arXiv preprint arXiv:1707.06347}, 2017.

\bibitem{wang2016dueling}
Wang, Z., Schaul, T., Hessel, M., Van Hasselt, H., Lanctot, M., \&
  Freitas, N.
\newblock Dueling network architectures for deep reinforcement learning.
\newblock In {\it Proceedings of the 33rd International Conference on Machine
  Learning (ICML)}, 2016.

\bibitem{vanhasselt2016deep}
Van Hasselt, H., Guez, A., \& Silver, D.
\newblock Deep reinforcement learning with double Q-learning.
\newblock In {\it Proceedings of the 30th AAAI Conference on Artificial
  Intelligence}, 2016.

\bibitem{mnih2013playing}
Mnih, V., Kavukcuoglu, K., Silver, D., Graves, A., Antonoglou, I.,
  Wierstra, D., \& Riedmiller, M.
\newblock Playing Atari with deep reinforcement learning.
\newblock {\it arXiv preprint arXiv:1312.5602}, 2013.

\end{thebibliography}
}


% Appendix
\appendix
\section{Reproducibility Instructions}

All commands assume execution from the repository root directory
(e.g., \texttt{/path/to/work}).

\subsection{Validation Harness}
\begin{verbatim}
# Smoke tests (22 tests, no GPU required)
PYTHONPATH=$PWD/coder:$PYTHONPATH python coder/smoke_test.py

# Comprehensive test suite (53 tests)
PYTHONPATH=$PWD/coder:$PYTHONPATH pytest validator/test_model.py -v --tb=short
\end{verbatim}

\subsection{Ablation Study}
\begin{verbatim}
PYTHONPATH=$PWD/coder:$PYTHONPATH python validator/run_ablations.py
\end{verbatim}

\subsection{Profiling}
\begin{verbatim}
# Forward pass profile
PYTHONPATH=$PWD/coder:$PYTHONPATH \
  python validator/profile_model.py --mode forward --steps 20

# Training step profile
PYTHONPATH=$PWD/coder:$PYTHONPATH \
  python validator/profile_model.py --mode train --steps 10

# Action expansion profile
PYTHONPATH=$PWD/coder:$PYTHONPATH \
  python validator/profile_model.py --mode expansion

# Component parameter counts
PYTHONPATH=$PWD/coder:$PYTHONPATH \
  python validator/profile_model.py --mode component_params
\end{verbatim}

\subsection{Demoware (CartPole with Simulated Expansion)}
\begin{verbatim}
PYTHONPATH=$PWD/coder:$PYTHONPATH \
  python coder/train.py --env cartpole --n_episodes 2000
\end{verbatim}

\subsection{Dependencies}
\begin{itemize}
  \item Python $\geq$ 3.9, PyTorch $\geq$ 2.0, Gymnasium $\geq$ 0.28, NumPy
  \item No GPU required for any test, ablation, or profiling command
  \item All hyperparameters are documented in \texttt{coder/model\_config.py}
  as a typed \texttt{ModelConfig} dataclass with sensible defaults
\end{itemize}


% NeurIPS Checklist
\newpage
\section*{NeurIPS Paper Checklist}

\paragraph{1. Claims.}
Do the main claims made in the abstract and introduction accurately reflect the
paper's contributions and scope?
\textbf{Yes.} The paper claims only to present a reference architecture and
validation harness for the problem of unexpected action-space expansion. All
contributions are phrased in terms of design artifacts and validated structural
properties, not empirical superiority. The absence of trained-model evaluation
and real-task results is explicitly stated throughout.

\paragraph{2. Experiments.}
Does the paper provide sufficient information to reproduce the experiments?
\textbf{Yes.} The validation harness produces deterministic results with a
fixed seed. Full reproducibility instructions, including exact commands for all
53 tests, the 14-condition ablation runner, and profiling scripts, are provided
in the appendix. All hyperparameters are documented in the
\texttt{ModelConfig} dataclass with defaults listed in Table~\ref{tab:config}.

\paragraph{3. Theory.}
Does the paper make theoretical contributions and, if so, are they supported?
\textbf{Partially.} The paper leverages the monotonicity property of nested
action spaces~\cite{farquhar2020growing} to justify the TD target formulation.
No new theoretical results (theorems, proofs) are presented; the $k$-NN
centroid initialisation is motivated as a heuristic rather than a formally
optimal choice.

\paragraph{4. Datasets.}
Does the paper use datasets and, if so, are they documented?
\textbf{No.} The paper does not use any real-world datasets. All evaluations
use synthetic random states or the CartPole-v1 environment from Gymnasium (for
demoware purposes only).

\paragraph{5. Code.}
Does the paper release code and, if so, is it documented?
\textbf{Yes.} The reference implementation is included in the accompanying
submission package. It includes:
\begin{itemize}
  \item \texttt{coder/model\_config.py}: typed dataclass for all hyperparameters
  \item \texttt{coder/networks.py}: all network components with docstrings
  \item \texttt{coder/agent.py}: \texttt{ActionIncrementalDQN} agent
  \item \texttt{coder/env\_wrapper.py}: \texttt{ExpandingActionWrapper}
  \item \texttt{coder/train.py}: training loop with CLI interface
  \item \texttt{coder/smoke\_test.py}: 22-test smoke suite
  \item \texttt{validator/test\_model.py}: 53-test pytest suite
  \item \texttt{validator/run\_ablations.py}: 14-condition ablation runner
  \item \texttt{validator/profile\_model.py}: profiling scripts
\end{itemize}

\paragraph{6. Broader impacts.}
Does the paper discuss potential broader impacts?
\textbf{Minimal.} This paper presents a reference architecture and validation
harness for a foundational RL capability---adapting to changing action spaces.
The work is at an early, pre-empirical stage. Potential societal impacts are
commensurate with those of deep RL generally and are not amplified by this
specific contribution. No immediate dual-use concerns are identified.

\end{document}
